The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-12-19 smurf_paper.tex
updates to model sections

In addition to specifying when each model component is calculated
(normally an attempt is made to order them by decreasing amplitude in
the time-series), there are a number of parameters that control their
operation. Table~\ref{tab:components} list the model components and
the typical order in which they are calculated. The following sections
describe their implementation in detail.

\begin{table}
  \caption{Summary of the model components that can be fit to
    \scuba\ time-series data with SMURF. Only the first group of
    models are typically fit to the data (\model{COM}--\model{NOI})
    in the indicated order. The remaining models
    (\model{DKS}--\model{PLN}) have only been included for
    completeness}
  \vspace{0.2cm}
  \centering
  \begin{tabular}{c|l}
    \hline
    Model & Description \\
    \hline
    \model{COM} & remove common-mode signal \\
    \model{GAI} & common-mode scaled to each bolometer \\
    \model{EXT} & extinction correction \\
    \model{FLT} & Fourier transform filter \\
    \model{AST} & map estimate of astronomical signal \\
    \model{NOI} & noise estimation \\
    \hline
    \model{DKS} & dark squid cleaning along columns \\
    \model{SMO} & time-domain smoothing filter \\
    \model{PLN} & 2-dimensional time-varying plane removal \\
    \hline
    \end{tabular}
  \label{tab:components}
\end{table}

@@

﻿Fig.~\ref{fig:com} shows the time series from a selection of typical
bolometers.\footnote{The data have been flat-fielded and each time series
has been adjusted to a mean value of zero.} The similarity between most
bolometers is evident, and forms the common-mode signal - assumed to be a
consequence of variations in the atmospheric emission and fridge
temperature. This common-mode usually dominates the astronomical signal
for all but the brightest sources, and swamps faint extended structure.
The purpose of the COM and GAI models is to remove this common-mode
signal.

@@

The COM model is the common-mode signal itself. It is a single time
series that is estimated by finding the mean of all bolometers at each
time step. Bolometer values that have been flagged as unusable for any
reason are excluded from the mean.

Since bolometers have varying sensitivities, the amplitude of the
common-mode variations will also vary from bolometer to bolometer. So
comparing each bolometer time stream with the common-mode signal allows
an estimate of the bolometer sensitivity to be obtained. In practice, a
least squares linear fit is performed between the bolometer time series
and the common-mode to determine a gain and offset for each bolometer.
The gain and offset for each bolometer is known as the GAI model. Each
bolometer value can then optionally be scaled and shifted using these
values so that all bolometers share a common (but as yet unknown)
calibration. This provides an alternative( or additional) flatfielding
strategy to that described in in section~\ref{sec:bolos}.

@@

Any extended astronomical emission on a scale comparable to or larger
than the spatial extent of the area used to estimate the common-mode will
contribute a similar signal to all bolometers. Therefore such extended
emission is indistinguishable from the other sources of common-mode
signal (fridge and atmosphere variations) and will be removed by the COM
model. This places a limit on the spatial extent of astronomical
structure that can be recovered.

For this reason, the usual practice is to estimate a single COM model by
examining data from all four sub-arrays in each waveband, since this
allow spatial structure on the scale of the whole focal plane to be
recovered. However, sometimes there is evidence that the common-mode
differs from one array to another, and so an option exists to estimate a
separate COM model for each individual subarray, with a consequent
lowering in the scale of spatial structure that can be recovered.

@@

The extinction correction is normally derived using the JCMT water
vapour radiometer (WVM), and is not considered to be a free parameter
in the solution. For details on how it is calculated see
\citet{dempsey2011}. Note that its numerical value is calculated only
once, and simply applied as a scale factor in the iterative solution.

@@

This model takes the FFT of the bolometer data, and can apply both
high- and low-pass filters, as well as notch filteres, at hard
frequency edges specified by the user. Alternatively, the frequency
edges of the filters may be defined in terms of an \emph{angular
  scale}, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series may optionally have apodization
applied before the transform to avoid ringing (primarily caused by
wrap-around discontinuities at the ends of the time-series). Finally,
a whitening filter may also be applied in which a simple $1/f^\alpha +
\mathrm{constant}$ is fit to the power spectrum of each bolometer, and
then the bolometer FFT is multiplied by its inverse (though normalized
to preserve the white noise level).  The signal that is \emph{removed}
from the time-series by this process are stored in the \model{FLT}
container array.

@@

Map estimation is accomplished using a simple nearest-neighbour
resampling of the data onto a pre-defined map grid. For the $i$th map
pixel $\mathbf{m}(x_i,y_i)$, the brightness is estimated as the
weighted average of the bolometer data samples $\mathbf{b}_j$ that
land within that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (i.e., low-frequency)
noise.

In addition, a variance map $\mathbf{v}(x_j,y_j)$ is estimated. The
default behaviour is to estimate this weighted error on the mean given
the scatter in the weighted samples. This is accomplished by dividing
the \emph{biased weighted sample variance} by the number of samples
that went into the average (akin to the formula for standard error on
the mean, but accounting for weights),
%
\begin{equation}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. We decided not to use the \emph{unbiased} estimator since in
practice it would require accumulating an additional array of values
at every map pixel, and only results in a small difference where there
are less than $\sim10$ samples per pixel (a situation that is almost
never encountered in a SCUBA-2 map except in the edge pixels).

@@

\subsubsection{Down-sampling}

>>>>

In addition to specifying when each model component is calculated
(normally an attempt is made to order them by decreasing amplitude in
the time-series), there are a number of parameters that control their
operation. Table~\ref{tab:components} lists the model components and
the typical order in which they are calculated. The following sections
describe their implementation in detail.

\begin{table}
  \caption{Summary of the model components that can be fit to
    \scuba\ time-series data with SMURF. Only the first group of
    models (in the order indicated) are typically fit to the data
    (\model{COM}--\model{NOI}) in the indicated order. The remaining
    models (\model{DKS}--\model{PLN}) have only been included for
    completeness}
  \vspace{0.2cm}
  \centering
  \begin{tabular}{c|l}
    \hline
    Model & Description \\
    \hline
    \model{COM} & remove common-mode signal \\
    \model{GAI} & common-mode scaled to each bolometer \\
    \model{EXT} & extinction correction \\
    \model{FLT} & Fourier transform filter \\
    \model{AST} & map estimate of astronomical signal \\
    \model{NOI} & noise estimation \\
    \hline
    \model{DKS} & dark squid cleaning along columns \\
    \model{SMO} & time-domain smoothing filter \\
    \model{PLN} & 2-dimensional time-varying plane removal \\
    \hline
    \end{tabular}
  \label{tab:components}
\end{table}

@@

﻿Fig.~\ref{fig:com} shows the time series from a selection of typical
bolometers.\footnote{The data have been flat-fielded and each time series
has been adjusted to a mean value of zero.} The similarity between most
bolometers is evident, and forms the common-mode signal - assumed to be a
consequence of variations in the atmospheric emission and fridge
temperature. This common-mode usually dominates the astronomical signal
for all but the brightest sources, and swamps faint extended structure.
The purpose of the \model{COM} and \model{GAI} models is to remove this common-mode
signal.

@@

The \model{COM} model is the common-mode signal itself. It is a single time
series that is estimated by finding the mean of all bolometers at each
time step. Bolometer values that have been flagged as unusable for any
reason are excluded from the mean.

Since bolometers have varying sensitivities, the amplitude of the
common-mode variations will also vary from bolometer to bolometer. So
comparing each bolometer time stream with the common-mode signal
allows an estimate of the bolometer sensitivity to be obtained. In
practice, a least squares linear fit is performed between the
bolometer time series and the common-mode to determine a gain and
offset for each bolometer.  The gain and offset for each bolometer is
known as the \model{GAI} model. Each bolometer value can then
optionally be scaled and shifted using these values so that all
bolometers share a common (but as yet unknown) calibration. This
provides an alternative( or additional) flatfielding strategy to that
described in in section~\ref{sec:bolos}.

@@

Any extended astronomical emission on a scale comparable to or larger
than the spatial extent of the area used to estimate the common-mode will
contribute a similar signal to all bolometers. Therefore such extended
emission is indistinguishable from the other sources of common-mode
signal (fridge and atmosphere variations) and will be removed by the 
model. This places a limit on the spatial extent of astronomical
structure that can be recovered.

For this reason, the usual practice is to estimate a single
\model{COM} model by examining data from all four sub-arrays in each
waveband, since this allow spatial structure on the scale of the whole
focal plane to be recovered. However, sometimes there is evidence that
the common-mode differs from one array to another, and so an option
exists to estimate a separate \model{COM} model for each individual
subarray, with a consequent lowering in the scale of spatial structure
that can be recovered.

@@

The extinction correction is a multiplicative factor that is normally
derived using the JCMT water vapour radiometer (WVM), and is not
considered to be a free parameter in the solution. However, it is
applied as part of the iterative solution, rather than a
pre-processing step, since any small errors will be amplified by the
large low-frequency drifts in the raw bolometer time-series. For
example, if the bolometer drift is 1000 times greater than an
astronomical source of interest, a 1 per cent error in the flatfield
will produce stripes of order 10 times the astronomical signal
amplitude in the final map! If \model{EXT} is applied after the bulk
of the low-frequency noise has been removed (e.g., \model{COM,GAI}),
then there is little potential for such small errors to affect the
final map.  For details on how it is calculated see
\citet{dempsey2011}. Note that its numerical value is calculated only
once, and simply applied as a scale factor in the iterative solution.

@@

This model takes the FFT of the bolometer data, and can apply both
high- and low-pass filters, as well as notch filteres, at hard
frequency edges specified by the user. Alternatively, the frequency
edges of the filters may be defined in terms of an \emph{angular
  scale}, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series may optionally have apodization
applied before the transform to avoid ringing (primarily caused by
wrap-around discontinuities at the ends of the time-series). Finally,
a whitening filter may also be applied in which a simple $1/f^\alpha +
\mathrm{constant}$ is fit to the power spectrum of each bolometer, and
then the bolometer FFT is multiplied by its inverse (though normalized
to preserve the white noise level).  The signal that is \emph{removed}
from the time-series by this process are stored in the \model{FLT}
container array. Typically this model is used purely as a high-pass
filter to remove most of the $1/f$ noise. Low-pass filtering is
redundant for two reasons: (i) SMURF already low-pass filters and
re-samples to a lower sample rate as described in
Section~\ref{sec:downsamp}; and (ii) the act of re-gridding the data
to produce map estimates effectively low-pass filters the data at a
frequency that corresponds to the inverse of the crossing time of a
single map pixel. Notch filters have not been used with \scuba\ data
particularly since line features tend to move around (i.e. a dynamic
line-detection system would have to be developed, and care would have
to be taken that astronomical sources are not suppressed).

@@

Map estimation is accomplished using a nearest-neighbour resampling of
the data onto a pre-defined map grid. For the $i$th map pixel
$\mathbf{m}(x_i,y_i)$, the brightness is estimated as the weighted
average of the bolometer data samples $\mathbf{b}_j$ that land within
that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (i.e., low-frequency)
noise.

In addition, a variance map $\mathbf{v}(x_j,y_j)$ is estimated. The
default behaviour is to estimate this weighted error on the mean given
the scatter in the weighted samples. This is accomplished by dividing
the \emph{biased weighted sample variance} by the number of samples
that went into the average (akin to the formula for standard error on
the mean, but accounting for weights),
%
\begin{equation}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

 We decided not to use the \emph{unbiased} estimator since in
practice it would require accumulating an additional array of values
at every map pixel, and only results in a small difference where there
are less than $\sim10$ samples per pixel (a situation that is almost
never encountered in a SCUBA-2 map except in the edge pixels).

@@

\subsubsection{Down-sampling}
\label{sec:downsamp}
